---
title: "Lab 2 - Descriptive Statistics and Basic Mapping"
author: "GIS III Spring 2020 - Erin Abbott"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---
Submission by Sunday 4/19



## Data Manipulation

* Load Libraries
```{r, echo=TRUE}
library(sf)
library(raster)
library(dplyr)
library(spData)
library(tidyverse)
library(RColorBrewer)
library(ggplot2)

```

* Data sources:
  + Coffee production dataset: by  Roger Bivand, Jakub Nowosad, Robin Lovelace. Available on the [GitHub page](https://github.com/Nowosad/spData/tree/master/data).

  + World dataset: by  Roger Bivand, Jakub Nowosad, Robin Lovelace. Available on the [GitHub page](https://github.com/Nowosad/spData/tree/master/data).



```{r}
#load coffee production data
cof_data <- coffee_data
class(cof_data)

#load world data
world <- world
class(world)

#join coffee and world data
coffee <- left_join(world, cof_data)
head(coffee)

# create coffee production difference variable
coffee1 <- coffee %>% mutate(prod_dif = coffee_production_2017 - coffee_production_2016)
coffee1

```

 * Summary Stats:
```{r}
summary(coffee1$coffee_production_2016)
summary(coffee1$coffee_production_2017)
summary(coffee1$prod_dif)
summary(coffee$area_km2)
```
 
 
## Non-Spatial Data Visualization

```{r}
ggplot(data=coffee1,aes(prod_dif)) +
  geom_histogram(aes(fill=continent), color="black",bins=7) +
  theme_light() + 
  scale_fill_brewer(palette="Set3") +
  ggtitle("Change in Coffee Production between 2016 and 2017 by Continent") +
  xlab("Production Difference in (thousand 60-Kg Bags)") +
  ylab("Number of Countries") + 
  theme(plot.title = element_text(face = "bold"))


```



## Spatial Data Visualization

```{r}
plot(coffee1["prod_dif"], main="Change in Coffee Production between 2016 and 2017\n (in thousands of 60-Kg bags)", cex.main=0.8, axes = TRUE)

```

* Rendering in html

* Uploaded to GitHub